---
title: "Data Science for Statisticians: Introduction to the tidyverse"
author: "Rafael A. Irizarry"
date: "3/13/2019"
output: 
  ioslides_presentation: 
    fig_caption: no
    fig_height: 4
    fig_width: 7
    beamer_presentation: default
  slidy_presentation: default
---

## The tidyverse

A collection of packages particularly helpful for working with tables.


- __dplyr__ - manipulating data frames 
- __ggplot2__ - data visualization
- __readr__ - importing data
- __stringr__ - string manipulation
- __lubridate__ - dates and times
- __purrr__ package for working with functions


## The tidyverse

Load the main packages like this:

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
```


## Tidy data 

- Each row represents one observation 
- Columns represent the different variables 


## Tidy data 

The `murders` dataset is an example of a tidy data frame.

```{r}
library(dslabs)
data(murders)
head(murders)
```

## Wide data

This is not tidy:

```{r,  echo = FALSE, message=FALSE}
path <- system.file("extdata", package="dslabs")
filename <- file.path(path, "fertility-two-countries-example.csv")
wide_data <- read_csv(filename)
select(wide_data, country, `1960`:`1965`) %>% as.data.frame
```

Later we learn how to _wrangle_ into tidy format.

## Tidy data

Same information in tidy format. 

```{r, echo=FALSE}
library(dslabs)
data("gapminder")
tidy_data <- gapminder %>% 
  filter(country %in% c("South Korea", "Germany") & !is.na(fertility)) %>%
  select(country, year, fertility)
head(tidy_data, 12)
```


## Manipulating data frames with dplyr

- `mutate` - change or add columns   
- `filter`-  filter rows based on a rule
- `select` - Subset the data by selecting specific columns

## Adding a column with `mutate`


```{r, message=FALSE}
data("murders")
murders <- mutate(murders, rate = total / population * 100000)
```

Notice that here we used `total` and `population` inside the function, which are objects that are **not** defined in our workspace. 

## Adding a column with `mutate`

We can see the new column now

```{r}
head(murders)
```


## Subsetting with `filter`

```{r}
filter(murders, rate <= 0.71)
```


## Selecting columns with `select`

```{r}
new_table <- select(murders, state, region, rate)
filter(new_table, rate <= 0.71)
```

## The pipe: `%>%`

$$ \mbox{original data }
\rightarrow \mbox{ select }
\rightarrow \mbox{ filter } $$


## The pipe: `%>%`


When using the pipe, functions take left hand side as first argumemt:

```{r}
murders %>% select(state, region, rate) %>% filter(rate <= 0.71)
```

## More pipe example


```{r}
sqrt(16)
16 %>% sqrt()
```

```{r}
log2(sqrt(16))
16 %>% sqrt() %>% log2()
```

## Example: Use arguments with pipe
```{r}
16 %>% sqrt() %>% log(base = 2)
```


## Pipe example

```{r, eval=FALSE}
murders %>% 
  select(state, region, rate) %>% 
  filter(rate <= 0.71)
```

## Summarizing data

Summarizing variables is very common operation.

Average female heights:

```{r}
data(heights)
s <- heights %>% 
  filter(sex == "Female") %>%
  summarize(average = mean(height))
```


## Tidyverse always returns tables

```{r}
s$average
```

## Summarizing data

You can compute more than one summary

```{r}
data(heights)
heights %>% 
  filter(sex == "Female") %>%
  summarize(median = median(height), 
            q25 = quantile(height, .25),
            q75 = quantile(height, .75))
```


## Summarize only wors with one variable summaries

You can compute more than one summary

```{r, eval=FALSE}
data(heights)
heights %>% 
  filter(sex == "Female") %>%
  summarize(qs = quantile(height, c(1,2,3) / 4))
```

```
Error in summarise_impl(.data, dots) : 
  Column `qs` must be length 1 (a summary value), not 3
```


## `pull`

The `us_murder_rate` object defined above represents just one number. Yet we are storing it in a data frame:

```{r}
s <- heights %>% 
  filter(sex == "Female") %>%
  summarize(average = mean(height))
class(s)
```

## `pull`

```{r}
s %>% pull(average)
```

## Group then summarize


```{r}
heights %>% group_by(sex)
```

- The result does not look very different from `heights`, except we see `Groups: sex [2]` when we print the object. 
- Although not immediately obvious from its appearance, this is now a special data frame called a _grouped data frame_ 


## Group then summarize

```{r}
heights %>% 
  group_by(sex) %>%
  summarize(average = mean(height), sd = sd(height))
```

## Sorting data frames


```{r}
murders %>%
  arrange(population) %>%
  head()
```

## Sorting data frames

```{r}
murders %>% 
  arrange(rate) %>% 
  head()
```

## Sorting data frames

```{r}
murders %>% 
  arrange(desc(rate)) %>% 
  head()
```

## Nested sorting 

```{r}
murders %>% 
  arrange(region, rate) %>% 
  head()
```


## The top $n$ 


```{r}
murders %>% top_n(10, rate)
```


## Tibbles


```{r}
data(murders)
class(murders)
```

- Remember `group_by`  permits stratifying data before computing summary statistics. 
- But where is the group information stored in the data frame? 

## Tibbles


```{r}
murders %>% group_by(region) %>% head()
```

## Tibbles

```{r}
murders %>% group_by(region) %>% class()
```

## Tibles versus data frames

- Tibbles display better 
- Subsets of tibbles are tibbles

```{r}
class(murders[,4])
class(as_tibble(murders)[,4])
```

## Tibles versus data frames

```{r}
murders$Population
as_tibble(murders)$Population
```


## Tibles versus data frames

- Tibbles can have complex entries

```{r}
tibble(id = c(1, 2, 3), func = c(mean, median, sd))
```

## Tibles versus data frames

- Tibbles can be grouped

## Create a tibble using `data_frame` 

```{r}
grades <- data_frame(names = c("John", "Juan", "Jean", "Yao"), 
                     exam_1 = c(95, 80, 90, 85), 
                     exam_2 = c(90, 85, 85, 90))
grades
```

## Another difference

```{r}
grades <- data.frame(names = c("John", "Juan", "Jean", "Yao"), 
                     exam_1 = c(95, 80, 90, 85), 
                     exam_2 = c(90, 85, 85, 90))
class(grades$names)
```

## Another difference

- To avoid this, we use the rather cumbersome argument `stringsAsFactors`:
```{r}
grades <- data.frame(names = c("John", "Juan", "Jean", "Yao"), 
                     exam_1 = c(95, 80, 90, 85), 
                     exam_2 = c(90, 85, 85, 90),
                     stringsAsFactors = FALSE)
class(grades$names)
```

## Conver to tibble

```{r}
as_tibble(grades) %>% class()
```

