---
title: "Introduction to ggplot2"
author: "Rafael A. Irizarry"
date: "17-3-7"
output:
  ioslides_presentation:
    fig_caption: no
    fig_height: 4
    fig_width: 7
  beamer_presentation: default
  slidy_presentation: default
---


## ggplot2 part of tidyverse

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
```

or


```{r, message=FALSE, warning=FALSE}
library(tidyverse)
```


## Other alternatives

- R base
- __grid__ 
- __lattice__. 


## Grammar of graphics 

__ggplot2__ a [grammar of graphics](http://www.springer.com/us/book/9780387245447), 

- Construct hundreds of different sentences by learning just a handful of verbs, nouns and adjectives without having to memorize each specific sentence. 


## Strength and limitation

- __ggplot2__ is easier for beginners is that its default behavior is carefully chosen to satisfy the great majority of cases and is visually pleasing.

- One limitation is that __ggplot2__ is designed to work exclusively with data tables in which rows are observations and columns are variables. 

## Sheet cheats 


- You should have the [ggplot2 sheet cheat](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf) handy. 

- To find perform an internet search for "ggplot2 cheat sheet".

## The components of a graph


```{r ggplot-example-plot, echo=FALSE}
library(dslabs)
data(murders)
library(ggthemes)
library(ggrepel)

r <- murders %>% 
  summarize(pop=sum(population), tot=sum(total)) %>% 
  mutate(rate = tot/pop*10^6) %>% pull(rate)

murders %>% ggplot(aes(x = population/10^6, y = total, label = abb)) +  
  geom_abline(intercept = log10(r), lty=2, col="darkgrey") +
  geom_point(aes(color=region), size = 3) +
  geom_text_repel() + 
  scale_x_log10() +
  scale_y_log10() +
  xlab("Populations in millions (log scale)") + 
  ylab("Total number of murders (log scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name="Region") +
  theme_economist()
```


## Components

* __Data__: The US murders data table is being summarized. We refer to this as the __data__ component. 

*  __Geometry__: The plot above is a scatterplot. This is referred to as the 
__geometry__ component. 

* __Aesthetic mapping__: The plot uses several visual cues to represent the information provided by the dataset. T


## Other minor components

* The points are labeled with the state abbreviations.
* The range of the x-axis and y-axis appears to be defined by the range of the data. They are both on log-scales. 
* There are labels, a title, a legend, and we use the style of The Economist magazine.

## Load data

```{r}
library(dslabs)
data(murders)
```

## `ggplot` objects 

```{r ggplot-example-1, eval=FALSE}
ggplot(data = murders)
```

## `ggplot` objects 

```{r ggplot-example-2}
murders %>% ggplot()
```


## `ggplot` objects 

```{r}
p <- ggplot(data = murders)
class(p)
```

## Rendering a plot

```{r, eval=FALSE}
print(p)
p
```

## Geometries

>> DATA %>% `ggplot()` + LAYER 1 + LAYER 2 + ... + LAYER N


## Geometries

- Geometry function names follow the pattern: `geom_X` 
- where X is the name of the geometry. 
- Examples include `geom_point`, `geom_bar` and `geom_histogram`.


## Aesthetics for geometries

- Look at help file. Here is `geom_point`

```
> Aesthetics
> 
> geom_point understands the following aesthetics (required aesthetics are in bold):
>
> x
>
> y
> 
> alpha
>
> colour
```


## Aesthetic mappings
 

```{r, eval = FALSE}
murders %>% ggplot() + 
  geom_point(aes(x = population/10^6, y = total))
```
 
- We can drop the `x = ` and `y =` if we wanted to since these are the first and second expected arguments, as seen in the help page. 

## We can add to an existing plot object

```{r ggplot-example-3}
p + geom_point(aes(population/10^6, total))
```


## Layers

```{r  ggplot-example-4}
p + geom_point(aes(population/10^6, total)) +
  geom_text(aes(population/10^6, total, label = abb))
```

## Where are variabled defined?

This is fine:

```{r, eval=FALSE}
p_test <- p + geom_text(aes(population/10^6, total, label = abb))
```

## Where are variabled defined?

This is not:
```{r, eval=FALSE}
p_test <- p + geom_text(aes(population/10^6, total), label = abb) 
```
## Tinkering with arguments
 
```{r ggplot-example-5}
p + geom_point(aes(population/10^6, total), size = 3) +
  geom_text(aes(population/10^6, total, label = abb))
```


## Tinkering with arguments

```{r ggplot-example-6}
p + geom_point(aes(population/10^6, total), size = 3) +
  geom_text(aes(population/10^6, total, label = abb), nudge_x = 1)
```

## Global versus local aesthetic mappings


```{r}
args(ggplot)
```


## Global versus local aesthetic mappings

```{r}
p <- murders %>% ggplot(aes(population/10^6, total, label = abb))
```

## Global versus local aesthetic mappings

```{r ggplot-example-7}
p + geom_point(size = 3) + 
  geom_text(nudge_x = 1.5)
```

## Global versus local aesthetic mappings

```{r ggplot-example-8}
p + geom_point(size = 3) +  
  geom_text(aes(x = 10, y = 800, label = "Hello there!"))
```


## Scales

```{r ggplot-example-9}
p + geom_point(size = 3) + geom_text(nudge_x = 0.05) + 
  scale_x_continuous(trans = "log10") +
  scale_y_continuous(trans = "log10") 
```

## Scales

```{r, eval=FALSE}
p + geom_point(size = 3) +  
  geom_text(nudge_x = 0.05) + 
  scale_x_log10() +
  scale_y_log10() 
```


## Labels and titles
```{r, eval=FALSE}
p + geom_point(size = 3) +  
  geom_text(nudge_x = 0.05) + 
  scale_x_log10() +
  scale_y_log10() +
  xlab("Populations in millions (log scale)") + 
  ylab("Total number of murders (log scale)") +
  ggtitle("US Gun Murders in 2010")
```

## Labels and titles
```{r ggplot-example-10, echo=FALSE}
p + geom_point(size = 3) +  
  geom_text(nudge_x = 0.05) + 
  scale_x_log10() +
  scale_y_log10() +
  xlab("Populations in millions (log scale)") + 
  ylab("Total number of murders (log scale)") +
  ggtitle("US Gun Murders in 2010")
```


## Categories as colors

```{r, eval=FALSE}
p <-  murders %>% ggplot(aes(population/10^6, total, label = abb)) +   
  geom_text(nudge_x = 0.05) + 
  scale_x_log10() +
  scale_y_log10() +
  xlab("Populations in millions (log scale)") + 
  ylab("Total number of murders (log scale)") +
  ggtitle("US Gun Murders in 2010")
```

## Categories as colors

This won't work
```{r, echo=FALSE}
p <-  murders %>% ggplot(aes(population/10^6, total, label = abb)) +   
  geom_text(nudge_x = 0.05) + 
  scale_x_log10() +
  scale_y_log10() +
  xlab("Populations in millions (log scale)") + 
  ylab("Total number of murders (log scale)") +
  ggtitle("US Gun Murders in 2010")
```

```{r ggplot-example-11}
p + geom_point(size = 3, color ="blue")
```

## Example: color as a mapping


```{r ggplot-example-12}
p + geom_point(aes(col=region), size = 3)
```

## Annotation, shapes, and adjustments

```{r}
r <- murders %>% 
  summarize(rate = sum(total) /  sum(population) * 10^6) %>% 
  pull(rate)
```


## Annotation, shapes, and adjustments

```{r ggplot-example-13}
p + geom_point(aes(col=region), size = 3) + 
  geom_abline(intercept = log10(r))
```

## Annotation, shapes, and adjustments

```{r}
p <- p + geom_abline(intercept = log10(r), lty = 2, color = "darkgrey") +
  geom_point(aes(col=region), size = 3)  
```

## Annotation, shapes, and adjustments

```{r}
p <- p + scale_color_discrete(name = "Region") 
```

## Add-on packages 

```{r}
library(ggthemes)
p + theme_economist()
```

## Another example

```{r}
library(ggthemes)
p + theme_fivethirtyeight()
```

## Putting it all together


```{r, eval=FALSE}
r <- murders %>% 
  summarize(rate = sum(total) /  sum(population) * 10^6) %>%
  pull(rate)

murders %>% ggplot(aes(population/10^6, total, label = abb)) +   
  geom_abline(intercept = log10(r), lty = 2, color = "darkgrey") +
  geom_point(aes(col=region), size = 3) +
  geom_text_repel() + 
  scale_x_log10() +
  scale_y_log10() +
  xlab("Populations in millions (log scale)") + 
  ylab("Total number of murders (log scale)") +
  ggtitle("US Gun Murders in 2010") + 
  scale_color_discrete(name = "Region") +
  theme_economist()
```


## Putting it all together


```{r final-ggplot-example, echo=FALSE}
library(ggthemes)
library(ggrepel)

r <- murders %>% 
  summarize(rate = sum(total) /  sum(population) * 10^6) %>%
  pull(rate)

murders %>% ggplot(aes(population/10^6, total, label = abb)) +   
  geom_abline(intercept = log10(r), lty = 2, color = "darkgrey") +
  geom_point(aes(col=region), size = 3) +
  geom_text_repel() + 
  scale_x_log10() +
  scale_y_log10() +
  xlab("Populations in millions (log scale)") + 
  ylab("Total number of murders (log scale)") +
  ggtitle("US Gun Murders in 2010") + 
  scale_color_discrete(name = "Region") +
  theme_economist()
```

## Quick plots with `qplot` {#qplot}

Make a quick scatterplot:

```{r fig.height=3, fig.width=4.5}
data(murders)
x <- log10(murders$population)
y <- murders$total
qplot(x, y)
```


## Grids of plots

There are often reasons to graph plots next to each other. The __gridExtra__ package permits us to do that:

```{r, warning=FALSE, message=FALSE, out.width='100%', fig.height=3}
library(gridExtra)
p1 <- murders %>%
  mutate(rate = total/population*10^5) %>%
  filter(population < 2*10^6) %>%
  ggplot(aes(population/10^6, rate, label = abb)) +
  geom_text() +
  ggtitle("Small States")
p2 <- murders %>%
  mutate(rate = total/population*10^5) %>%
  filter(population > 10*10^6) %>%
  ggplot(aes(population/10^6, rate, label = abb)) +
  geom_text() +
  ggtitle("Large States")
```

## Grids of plots

```{r grid-of-plots}
grid.arrange(p1, p2, ncol = 2)
```
